The Design and Implementation of the Synchronous Language \CEU

### 2016-04-04 paper.tex
...

\usepackage{verbatim}
\usepackage{xspace}
\newcommand{\C}{\emph{C}\xspace}
\newcommand{\VM}{\emph{VM}\xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\ax}{\code{[a]}\xspace}
\newcommand{\bx}{\code{[b]}\xspace}
\newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}

@@

% TODO
Nested \code{emit} invocations (e.g., suppose we \code{emit c} after the 
\code{await a} in line 8) results in fine-grained micro reactions stacked one 
on top of the other, all inside the same reaction to an external event.
%Internal events provide a fine-grained intra-reaction control mechanism.
For instance, it brings a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself is always ignored, 
given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring data stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, similarly to coroutines~\cite{lua.coroutines}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.

@@

Esterel is only deterministic with respect to external behavior: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages, such as 
\emph{SOL}~\cite{wsn.sol},
\emph{SC}~\cite{rp.synchc}, and
\emph{PRET-C}~\cite{rp.pretc},
enforce intra-reaction determinism with an arbitrary execution order for 
statements in multiple lines of execution.
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

@@

Note that such analysis is only possible due to the uniqueness of input events 
within reactions.
Otherwise, any two trail segments in parallel could be concurrent, even if they 
react to different input events.

@@

\CEU aims to offer a simpler semantics than Esterel with more determinism and 
fine-grained control over the program execution.
%
The list that follows summarizes the contributions of our design in this 
direction:
%
\begin{itemize}
%
\item Event-triggered notion of time bound to the semantics of the language.
Event-driven programming is popular in many domains, such as server and GUI 
development.
We believe that programmers are more familiar with dealing with events in 
isolation, which simplifies the reasoning about concurrency.
In addition, the uniqueness of external events is a prerequisite for the 
temporal analysis of \CEU.
%
\item Deterministic intra-reaction execution and communication.
Determinism in \CEU is all-inclusive and does not depend on additional levels 
of static analysis.
It encompasses the whole language, including memory accesses, system calls, and 
stack-based internal events.
Programmers can always figure out which statement executes next, making runtime 
analysis and debugging easier.
%
\item Static temporal analysis for concurrent statements.
Although execution is deterministic, the compiler of \CEU still advises about 
suspicious statements that can react concurrently to the same event, forcing 
the to serialize them accordingly.
%
\item Safe integration with \C.
\CEU enforces the programmer to deal with pointers representing resources 
through finalization clauses.
When dealing with concurrent system calls, programmers can provide the compiler 
with annotations to remove false positives from the temporal analysis or to 
allow non commutative concurrent behavior.
%
\item First-class timers with dedicated syntax and automatic synchronization.
Given the omnipresence of timers in embedded systems, a dedicated syntax can 
simplify the development and readability of programs.
Furthermore, automatic synchronization releases the programmer from the burden 
of adjusting timers in sequence and in parallel.
%
\end{itemize}

Our synchronous and deterministic approach also leads to some limitations as 
follows:
%
\begin{itemize}
\item Execution is purely reactive as a result of event-triggered reactions.
Since only occurrences of events can start reactions, programs cannot execute 
proactively in the absence of events.
\item Reactions must execute in bounded time due to the synchronous hypothesis.
As a synchronous language, \CEU requires that CPU times for reactions are 
negligible in comparison to the rate of incoming events.
\item Execution is sequential because of intra-reaction determinism.
The deterministic semantics of \CEU does not make parallelization easy (to be 
discussed in Section~\ref{sec.impl.single}).
\end{itemize}
%
Nonetheless, we advocate for a tractable synchronous reactive core that 
supports shared memory with deterministic execution.
%
To deal with the limitations, we recommend for memory-isolated parallelizable 
asynchronous primitives as language extensions, but which are not in the scope 
of this paper.

@@

The compilation process of a program in \CEU for the original \C back end is 
composed of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an \emph{abstract 
syntax tree (AST)};
the \emph{temporal analysis phase} detects inconsistencies in programs, such as 
unbounded loops and suspicious accesses to shared memory;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and packs it with platform-dependent functionality (e.g., system calls) and the 
runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.

In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
in comparison to hand-crafted event-driven code in \emph{nesC}~\cite{wsn.nesc} 
(a \C variant) available in a stable codebase%
\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}.
%
Figure~\ref{fig.eval} compares source size (\emph{tokens}), binary size 
(\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
network protocols and a radio driver.
The small increase in resource usage shows that \CEU fits the context of 
constrained embedded systems which is typically reserved to \C only.

@@

- TODO: no available Esterel compiler for the constrained target platform
    - we compare with C, which is a benchmark, technically unbeatable for 
      resource efficiency

In the sections that follow, we discuss the implementation highlights related 
to the peculiarities of \CEU:
the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
single-threaded dispatching (Section~\ref{sec.impl.single}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

The compile-time \emph{temporal analysis} phase detects inconsistencies in \CEU 
programs.
Here, we focus on the algorithm that detects suspicious concurrent statements, 
such as accesses to shared variables, as discussed in 
Section~\ref{sec.ceu.analysis}.

For each node representing a statement in the program AST, we keep the set of 
input events $I$ (for \emph{incoming}) that can lead to the execution of the 
node, and also the set of input events $O$ (for \emph{outgoing}) that can 
terminate the node.

A node inherits the set $I$ from its direct parent and calculates $O$ according 
to its type:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, \C calls, and declarations 
simply reproduce $O=I$, as they do not await;
%
\item An \code{await E} statement has $O=\{E\}$, where \code{E} is an input 
event (see internal events below).
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never terminates, i.e., never proceeds to the statement 
immediately following it (see also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$.
The first child inherits $I$ from its parent node, while the set $O$ for the 
whole node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all \code{break} statements' $O$ forms the set $O$ for a 
\code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\item For internal events, an \code{await} awakes from any input that leads to 
    any matching \code{emit} in a trail in parallel:
    \begin{itemize}
    \item An \code{emit e} terminates in the same reaction, having $O=I$.
    \item An \code{await e} has $O=I_{e1} \cup ... \cup I_{eN}$,
        where $e1...eN$ are \code{emit e} statements in trails in parallel.
    \end{itemize}
\end{itemize}

With all sets calculated, we take all pairs of nodes that perform side effects 
and are in parallel branches, comparing their sets $I$ for intersections.
For each pair, if the intersection is not the empty set, we mark both nodes as 
suspicious.

The example in Figure~\ref{lst.impl.ast}.a has a corresponding $AST$, in 
Figure~\ref{lst.impl.ast}.b, with the sets $I$ and $O$ for each node.
The event $.$ (dot) represents the ``boot'' reaction.
The assignments to \code{y} in parallel (ln. 5,8 in the code) have an empty 
intersection of $I$ (ln. 6,9 in the AST), hence, they do not conflict.
Note that although the accesses in ln. 5,11 in the code (ln. 6,11 in the AST) 
do have an intersection, they are not in parallel and are also safe.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\centering\small{\ax}
\end{minipage}
%
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\centering\small{\bx}
\end{minipage}
%
%\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

\CEU favors a fine-grained use of trails, being common to use trails that await 
a single event.
For this reason, \CEU does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout generated by the compiler.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

@@

Each line of execution in \CEU needs to carry associated data, such as which 
event it is awaiting and which code to execute when it awakes.
%
The compiler statically infers the maximum number of trails a program can have 
at the same time and creates a static vector to hold the runtime information 
about them.
%
Like normal variables, trails that cannot be active at the same time can share 
slots in the static memory vector.

@@

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds an associated stack level and is delayed until the 
scheduler runtime reaches that level again.
%
A \code{FINALIZE} trail represents a hanged finalization block which is only 
scheduled when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the minimum sequence number (\emph{seqno}) in which it can awake.
%
In concrete terms, a trail is represented by the following \code{struct}:

\begin{lstlisting}
struct trail_t {
    state_t evt;
    label_t lbl;
    union {
        unsigned char seqno;
        stack_t       stk;
    };
};
\end{lstlisting}

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two 
segments and requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The \code{seqno} could eventually overflow during execution (i.e., every 256 
reactions).
However, given that the scheduler traverses all trails on every reaction, it 
can adjust them to properly handle overflows (actually, 2 bits to hold the 
\code{seqno} is already enough).
%
The size of \code{stack\_t} depends on the maximum depth of nested emissions 
and is bounded to the maximum number of trails.
In the worst case, a trail emits an event that awakes another trail, which 
emits an event that awakes another trail, and so on.
The last trail cannot awake any trail, because they are all hanged in the 
\code{STACKED} state.

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with an overhead of 2\% in comparison to the 
original version in \C~\cite{ceu.sensys13}.

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in \ax, the compiler extracts the entry points and associated 
trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in \bx).
The state of trails translates to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

\begin{comment}
%\subsection{ Discussion }
\label{sec.ceu.discussion}

>>>>

\usepackage{verbatim}
\usepackage{xspace}
\newcommand{\C}{\emph{C}\xspace}
\newcommand{\VM}{\emph{VM}\xspace}
\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\ax}{\code{[a]}\xspace}
\newcommand{\bx}{\code{[b]}\xspace}
\newcommand{\cx}{\code{[c]}\xspace}
\newcommand{\MM}[1] {\textcircled{\tiny{\textsf{#1}}}}

@@

Internal events provide a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

\CEU also supports nested \code{emit} invocations for internal events.
For instance, the code that executes after the \code{await inc} (ln. 4) could 
\code{emit} another event, creating a new level in the stack.
The runtime stack represents fine-grained micro reactions, one on top of the 
other, all inside the same reaction to an external event.
%Internal events provide a fine-grained intra-reaction control mechanism.

@@

On the one hand, this form of subroutines has a significant limitation that it 
cannot express recursive calls: an \code{emit} to itself is always ignored, 
given that a running body cannot be awaiting itself.
%
On the other hand, this very same limitation brings some important safety 
properties to subroutines:
first, they are guaranteed to react in bounded time;
second, memory for locals is also bounded, not requiring data stacks.
%
Also, this form of subroutines can use the other primitives of \CEU, such as 
parallel compositions and the \code{await} statement.
In particular, they await keeping context information such as locals and the 
program counter, similarly to coroutines~\cite{lua.coroutines}.
%
In a previous work, we build on top of internal events other advanced control 
mechanisms, such as resumable exceptions and reactive 
variables~\cite{ceu.rem13}.
%In Section~\ref{sec.adv.excpt} we show how to use them to implement 
%exceptions.

@@

Esterel is only deterministic with respect to external behavior: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages enforce intra-reaction 
determinism with an arbitrary execution order for statements in multiple lines 
of execution (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

@@

Note that such analysis is only possible due to the uniqueness of input events 
within reactions.
Otherwise, any two trail segments in parallel could be concurrent, even if they 
react to different input events.
%
Note also that the analysis is completely optional and does not affect the 
semantics of the program.
% TODO-no: static analsysis also simpler
% highlight that is not only important to have a comptutally-sound algorithm, 
% but that something that programmers can understand

@@

\CEU aims to offer a simpler semantics than Esterel with more determinism and 
fine-grained control over the program execution.
%
The list that follows summarizes the contributions of our design in this 
direction:
%
\begin{itemize}
%
\item Event-triggered notion of time bound to the semantics of the language.
Event-driven programming is popular in many domains, such as server and GUI 
development.
We believe that programmers are more familiar with dealing with events in 
isolation, which simplifies the reasoning about concurrency.
In addition, the uniqueness of external events is a prerequisite for the 
temporal analysis of \CEU.
%
\item Deterministic intra-reaction execution and communication.
Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
levels of static analysis.
It encompasses the whole language, including memory accesses, system calls, and 
stack-based internal events.
Programmers can always figure out which statement executes next, making runtime 
analysis and debugging easier.
%
\item Static temporal analysis for concurrent statements.
Although execution is deterministic, the compiler of \CEU still advises about 
suspicious statements that can react concurrently to the same event, forcing 
the to serialize them accordingly.
%
\item Safe integration with \C.
\CEU enforces the programmer to deal with pointers representing resources 
through finalization clauses.
When dealing with concurrent system calls, programmers can provide the compiler 
with annotations to remove false positives from the temporal analysis or to 
allow non commutative concurrent behavior.
%
\item First-class timers with dedicated syntax and automatic synchronization.
Given the omnipresence of timers in embedded systems, a dedicated syntax can 
simplify the development and readability of programs.
Furthermore, automatic synchronization releases the programmer from the burden 
of adjusting timers in sequence and in parallel.
%
\end{itemize}

Our synchronous and deterministic approach also leads to some limitations as 
follows:
%
\begin{itemize}
\item Execution is purely reactive as result of event-triggered reactions.
Since only occurrences of events can start reactions, programs cannot execute 
proactively in the absence of events.
\item Reactions must execute in bounded time due to the synchronous hypothesis.
As a synchronous language, \CEU requires CPU times for reactions to be 
negligible in comparison to the rate of incoming events.
\item Execution is sequential because of intra-reaction determinism.
%We discuss some implementation considerations in 
%Section~\ref{sec.impl.single}.
The deterministic semantics of \CEU does not make implicit parallelization easy 
(to be discussed in Section~\ref{sec.impl.single}).
\end{itemize}
%
Nonetheless, we advocate for keeping a tractable synchronous reactive core 
which supports shared memory concurrency with deterministic execution.
%
To deal with the limitations above, we recommend for memory-isolated 
parallelizable asynchronous functionality as a separate extension to the 
synchronous core~\cite{esterel.crp} (which is not in the scope of this paper).

\begin{comment}
TODO-no: parallel
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
\end{comment}

@@

The compilation process of programs in \CEU is composed of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
tree (\emph{AST});
the \emph{temporal analysis phase} detects inconsistencies in programs, such as 
unbounded loops and suspicious accesses to shared memory;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and packs it with platform-dependent functionality (e.g., system calls) and the 
runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.

In a previous work~\cite{ceu.sensys13}, we evaluate the implementation of \CEU
in comparison to a hand-crafted event-driven code base%
\footnote{TinyOS repository: \url{http://github.com/tinyos/tinyos-release/}}
in \emph{nesC} (a \C variant)~\cite{wsn.nesc}.
%
Figure~\ref{fig.eval} compares source size (\emph{tokens}), binary size 
(\emph{ROM}), and memory usage (\emph{RAM}) for a number of standardized 
network protocols and a radio driver.
The small increase in resource usage shows that the gains in productivity and 
safety with \CEU make it a viable alternative to \C in the context of 
constrained embedded systems.

Unfortunately, most real applications in Esterel seems to be closed source.
In a future work, we plan to cooperate with groups that have more experience 
and access to real code bases in Esterel for a comprehensive comparison with 
Céu.
% TODO-no:
%Still, considering \C as an ``absolute benchmark'', we consider the small 
%overhead .  solid results

@@

In the sections that follow, we discuss the implementation of subsystems that 
are specific (or in some way important) to \CEU:
the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
single-threaded dispatching (Section~\ref{sec.impl.single}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

The compile-time temporal analysis phase detects inconsistencies in \CEU 
programs.
Here, we focus on the algorithm that detects suspicious concurrent statements, 
such as accesses to shared variables, as discussed in 
Section~\ref{sec.ceu.analysis}.

For each node representing a statement in the program AST, we keep the set of 
input events $I$ (for \emph{incoming}) that can lead to the execution of the 
node, and also the set of input events $O$ (for \emph{outgoing}) that can 
terminate the node.
As an example, for the single-statement program \code{await~A}, we have 
$I=\{boot\}$ and $O=\{A\}$.

A node inherits the set $I$ from its immediate parent and calculates $O$ 
according to its type, as follows:
%
\begin{itemize}
%
\item Nodes that represent expressions, assignments, \C calls, and declarations 
simply reproduce $O=I$, as they do not await;
%
\item An \code{await E} statement, where \code{E} is an external input event,
      has $O=\{E\}$ (see also internal events below).
%
\item A \code{break} statement has $O=\{\}$ as it escapes the innermost 
\code{loop} and never proceeds to the statement immediately following it (see 
also \code{loop} below);
%
\item A \emph{sequence node (;)} modifies each of its children to have 
$I_n=O_{n-1}$, except for $n=1$ which inherits $I$ from the parent node.
The set $O$ for the whole node is copied from its last child, i.e., $O=O_n$.
%
\item A \code{loop} node includes its body's $O$ on its own $I$ ($I=I \cup 
O_{body}$), as the loop is also reached from its own body.
The union of all $O$ from nested \code{break} statements forms the set $O$ for 
a \code{loop}.
%
\item An \code{if} node has $O=O_{true} \cup O_{false}$.
%
\item A parallel composition may terminate from any of its branches, hence $O = 
O_1 \cup ... \cup O_n$.
\item For internal events, an \code{await} awakes from any input that leads to 
    any matching \code{emit} in a trail in parallel:
    \begin{itemize}
    \item An \code{emit e} terminates in the same reaction, having $O=I$.
    \item An \code{await e} has $O=I_{e1} \cup ... \cup I_{eN}$,
        where $e1...eN$ are \code{emit e} statements in trails in parallel.
    \end{itemize}
\end{itemize}

With all sets calculated, we take all pairs of nodes that perform side effects 
and are in parallel branches, and compare their sets $I$ for intersections.
For each pair, if the intersection is not empty, we mark both nodes as 
suspicious.

The code in Figure~\ref{lst.impl.ast}.a has the $AST$ and sets $I$ and $O$ for 
each node in Figure~\ref{lst.impl.ast}.b.
The event $.$ (dot) represents the ``boot'' reaction.
The assignments to \code{y} in parallel (ln. 5,8 in the code) have an empty 
intersection of $I$ (ln. 6,9 in the AST), hence, they do not conflict.
Note that although the accesses in ln. 5,11 in the code (ln. 6,11 in the AST) 
do have an intersection, they are not in parallel branches and are also safe.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input void A, B;
var int y;
par/or do
  await A;
  y = 1;
with
  await B;
  y = 2;
end
await A;
y = 3;
\end{lstlisting}
\centering\small{\ax} A program in \CEU...
\end{minipage}
%
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
Stmts I={.} O={A}
    Dcl_y I={.} O={.}
    ParOr I={.} O={A,B}
        Stmts I={.} O={A}
            Await_A I={.} O={A}
            Set_y I={A} O={A}
        Stmts I={.} O={B}
            Await_B I={.} O={B}
            Set_y I={B} O={B}
    Await_A I={A,B} O={A}
    Set_y I={A} O={A}
\end{lstlisting}
\centering\small{\bx} ...with corresponding sets $I$ and $O$.
\end{minipage}
%
%\rule{14cm}{0.37pt}
\caption{ A program with a corresponding AST describing the sets $I$ and $O$.
%{\small %\textmd{
The program is safe because accesses to \code{y} in parallel have no 
intersections for $I$.
%}%}
\label{lst.impl.ast}
}
\end{figure}

@@

\CEU favors a fine-grained use of trails, being common to use trails that await 
a single event and terminate.
For this reason, \CEU does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at any given time.
A position in the memory may hold different data (with variable sizes) during 
runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with its corresponding 
memory layout.
%
Each variable is assigned a unique $id$ so that variables with the same name 
can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
\ax:3).
%
The \code{do-end} blocks in sequence (ln. \ax:2--7) are packed in a 
\code{union} (ln. \bx:2--3), given that their variables cannot be in scope at 
the same time, e.g., \code{MEM.a\_1} and \code{MEM.b\_2} (ln. \bx:2,3) can 
safely share the same memory slot.
%
The example also illustrates the presence of runtime flags (ln. \bx:4--7) 
related to the \code{par/and} termination (ln. \ax:8--12), which also reside in 
reusable slots in the static memory.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\centering\small{\ax} A program in \CEU...
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;

.
\end{lstlisting}
\centering\small{\bx} ...with corresponding memory layout
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout generated by the compiler.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

@@

Each line of execution in \CEU needs to carry associated data, such as which 
event it is currently awaiting and which code to execute when it awakes.
%
The compiler statically infers the maximum number of trails a program can have 
at the same time and creates a static vector to hold the runtime information 
about them.
%
Like normal variables, trails that cannot be active at the same time share 
slots in the static memory vector.

@@

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds an associated numeric stack level and is delayed 
until the scheduler runtime reaches that level again.
%
A \code{FINALIZE} trail represents a hanged finalization block which is only 
scheduled when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the minimum sequence number (\emph{seqno}) in which it can awake, 
to prevent from awaking in the same reaction it is reached.
%
In concrete terms, a trail is represented by the following \code{struct}:

\begin{lstlisting}
struct trail_t {
    state_t evt;
    label_t lbl;
    union {
        unsigned char seqno;  // if evt=EVT_*
        stack_t       stk;    // if evt=STACKED
    };
};
\end{lstlisting}

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two 
segments and requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The \code{seqno} could eventually overflow during execution (i.e., every 256 
reactions).
However, given that the scheduler traverses all trails on every reaction, it 
adjusts them to properly handle overflows (actually, 2 bits to hold the 
\code{seqno} is already enough).
%
The size of \code{stack\_t} depends on the maximum depth of nested emissions 
and is bounded to the maximum number of trails.
In the worst case, a trail emits an event that awakes another trail, which 
emits an event that awakes another trail, and so on;
the last trail cannot awake any trail, because they are all hanged in the 
\code{STACKED} state.

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with memory overhead of only 2\% in comparison 
to the original version in \C~\cite{ceu.sensys13}.

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in \ax, the compiler extracts the entry points and associated 
trails, e.g., the label \code{Awake\_e} will execute on \code{TRAIL-0} (ln.  
\ax:7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail in two segments with 
associated entry points.
%
The entry points translate to an \code{enum} in the generated code (ln.  
\bx:1--10).
The state of trails translates to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. \bx:12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. \bx:13), while all others are set to \code{INACTIVE} (ln. \bx:14).

@@
